---
title: "ggspavis package overview"
author: 
  - name: Lukas M. Weber
    affiliation: "Johns Hopkins Bloomberg School of Public Health"
package: ggspavis
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{ggspavis package overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

The `ggspavis` package contains a set of visualization functions for spatially resolved transcriptomics datasets.

Functions in this package are designed to work with the [SpatialExperiment](https://bioconductor.org/packages/SpatialExperiment) Bioconductor object class.


# Functions

Currently the package contains the following plotting functions:

- `?plotVisium`
- `?plotSpots`
- `?plotDimRed`


# Examples

We load some example datasets from the [STdata](https://github.com/lmweber/STdata) package, and create some example plots using the plotting functions from the `ggspavis` package.

```{r, message=FALSE}
library(SpatialExperiment)
library(STdata)
library(ggspavis)
```


## 10x Genomics Visium: mouse coronal brain section

```{r, message=FALSE}
# load data in SpatialExperiment format
spe <- load_data("Visium_mouseCoronal")
```

```{r, message=FALSE, out.width="60%"}
# add some values in 'colData' to show using color scale
colData(spe)$sum <- colSums(counts(spe))

# example plots
plotVisium(spe, fill = "sum")
plotVisium(spe, fill = "sum", palette = "red")
```


## 10x Genomics Visium: human brain (DLPFC)

```{r, message=FALSE}
# load data in SpatialExperiment format
spe <- load_data("Visium_humanDLPFC")
```

```{r, message=FALSE, out.width="52%"}
# example plots
plotSpots(spe)
```

```{r, message=FALSE, out.width="65%"}
plotSpots(spe, discrete = "ground_truth", palette = "libd_layer_colors")
```


## seqFISH: mouse embryo

```{r, message=FALSE}
# load data in SpatialExperiment format
spe <- load_data("seqFISH_mouseEmbryo")
```

```{r, message=FALSE, out.width="60%"}
# example plots
plotMolecules(spe, molecule = "Sox2")
```

